---
title: "Maternal Health Assignment"
author: "Clare-Allelua-Kahuma"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Importing Libraries

tidy verse - Loads a collection of R packages for data science, including ggplot2, dplyr, tidyr, readr, purrr, and more.
dplyr - Provides a grammar for data manipulation, making it easier to filter, arrange, summarize, and mutate data.

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)

```

### Importing a dataset

Using read csv because...
Header = TRUE to ensure the top column is considered as the label for columns 

```{r}
data <- read.csv("C:/Users/ckahuma/Downloads/Maternal Health Risk Data Set.csv", header = TRUE)

```

### Checking if my data loaded and exploring the data

```{r}
cat("------- Head of the data ---------\n")
print(head(data))

cat("\n------ Structure of the data ----------\n")
str(data)

cat("\n------- Summary of the data ---------\n")
print(summary(data))

cat("\n-------- Dimension of the data --------\n")
print(dim(data))

cat("\n-------- Number of rows --------\n")
print(nrow(data))

cat("\n-------- Number of columns --------\n")
print(ncol(data))

cat("\n-------- Column Names --------\n")
print(colnames(data))

cat("\n-------- The data type of the dataset --------\n")
print(class(data))

cat("\n-------- Class of the columns --------\n")
print(sapply(data, class))

cat("\n-------- Total missing data in columns --------\n")
print(colSums(is.na(data)))

cat("\n-------- Total missing data in dataset --------\n")
print(sum(is.na(data)))


```

### Group according to risk level and find the mean age per group

```{r}
data %>%
  group_by(RiskLevel) %>%
  summarise(mean_age = mean(Age)) %>%
  arrange(desc(mean_age))

```

### Median Per Group using Diastolic

```{r}
data %>%
  group_by(RiskLevel) %>%
  summarise(median_diastolic = median(DiastolicBP)) %>%
  arrange(desc(median_diastolic))

```

###### Fun fact: If low risk and mid risk both had 75, how did R choose what goes above? Alphabetical order?

R maintains the original order of these rows from the dataset. This is known as <a href="https://www.geeksforgeeks.org/stable-and-unstable-sorting-algorithms/" target="_blank"> "stable sorting" </a>—it preserves the relative order of equal values as they appeared in the dataset before sorting. 

You can also write some html in the R markdown file. Dr, is this good or bad practice?


### Creating a new variable: Systolic Rate 
```{r}
data %>%
  mutate(Systolic_Rate = SystolicBP / Age)

```

### Filter participants of Age < 40 and average risk systolic for each risk level


```{r}
data %>%
  filter(Age<=40) %>%
  group_by(RiskLevel) %>%
  summarise(AverageSystolic = mean(SystolicBP))


```

Why cant i filter before....Is this because the output before this (slightly higher) doesn't have age at this point?

### Plots - Box Plot, histogram

Continuous variable....

In data reshaping, there are two main formats:

Wide format: Each variable is represented by its own column.
Long format: Each variable is represented by two columns—one for the variable name (like SystolicBP, DiastolicBP, etc.) and one for the values associated with that variable.

```{r}
data_long <- data %>%
  select(Age, SystolicBP, DiastolicBP, BS, BodyTemp, HeartRate, RiskLevel) %>%
  pivot_longer(cols = c(SystolicBP, DiastolicBP, BS, BodyTemp, HeartRate),
               names_to = "Measurement", values_to = "Value")

  ggplot(data_long, aes(x = RiskLevel, y = Value)) +
  geom_boxplot() +
  facet_wrap(~ Measurement, scales = "free_y") +
  labs(title = "Box Plot of Measurements against Risk Level",
       y = "Value", x = "Risk Level") +
  theme_minimal()
  
```


Histogram plot

```{r}
# Reshape the data to long format
data_long <- data %>%
  select(Age, SystolicBP, DiastolicBP, BS, BodyTemp, HeartRate, RiskLevel) %>%
  pivot_longer(cols = c(Age, SystolicBP, DiastolicBP, BS, BodyTemp, HeartRate),
               names_to = "Measurement", values_to = "Value")

# Corrected histogram plot (x-axis should be continuous)
ggplot(data_long, aes(x = Value, fill = RiskLevel)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity") +
  facet_wrap(~ Measurement, scales = "free_x") +
  labs(title = "Histograms of Measurements", 
       x = "Value", y = "Count") +
  theme_minimal()
  

```

###Linear Regression

```{r}

heart_rate <- data$HeartRate
Age <- data$Age
plot(heart_rate, Age)

# lm start with the response variables and ~
fit <- lm(heart_rate~Age)
fit
summary(fit)

```

```{r}

heart_rate <- data$HeartRate
Systolic_BP <- data$SystolicBP
plot(heart_rate, Systolic_BP)

# lm start with the response variables and ~
fit <- lm(heart_rate~Systolic_BP)
fit
summary(fit)

```

```{r}

heart_rate <- data$HeartRate
Diastolic_BP <- data$DiastolicBP
plot(heart_rate, Diastolic_BP)

# lm start with the response variables and ~
fit <- lm(heart_rate~Diastolic_BP)
fit
summary(fit)

```


Check the predictors against themselves?

```{r}

library(corrplot)

# Compute correlation matrix
cor_matrix <- cor(data[, c("Age", "DiastolicBP", "SystolicBP")])

# Plot the correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.8)


```
Diastolic and systolic are hghly corelated
A inear model for  heart rate vs age,and systolic

```{r}

# Create a linear regression model
model <- lm(HeartRate ~ Age + Diastolic_BP, data = data)

# View model summary
summary(model)



```



es




















